\Ace: Active Typechecking and Translation Inside a Python

### 2014-05-10 ace-oopsla14/ace-oopsla14.tex
fixed some things

%\documentclass[12pt]{article}
\documentclass[10pt,preprint]{sigplanconf}
\newcommand{\lamAce}{\lambda_{\text{Ace}}}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
    \makeatletter
%
% \btIfInRange{number}{range list}{TRUE}{FALSE}
%
% Test if int number <number> is element of a (comma separated) list of ranges
% (such as: {1,3-5,7,10-12,14}) and processes <TRUE> or <FALSE> respectively
%
        \newcount\bt@rangea
        \newcount\bt@rangeb

@@

def print_transfer(t):
  print ("Transferring %s to %s." % (
    __ace_0.decimal_to_str(t[0],2), t[1][0]))

@@

\section{Theoretical Foundations}\label{theory}
\newcommand{\F}[1]{\textsf{#1}~}
\newcommand{\FF}[1]{\textsf{#1}}
\newcommand{\Q}{\FF{Arg}}
\renewcommand{\tnil}[1]{[]}
\begin{figure}[t]

@@

$$\begin{array}{rcl}	
%\textbf{programs} & 
\rho & ::= & \F{tycon}\fvar{tycon}~\F{of}\kappaidx~{\{}TC{\}}; \rho \pipe e\\ %\pfam{\familyDf}{\progsort}
%\\&  \pipe & \pdef{t}{\kappa}{\tau}{\progsort}  \pipe e
%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
 TC & ::= & \F{iana} \{\tau\}; \F{isyn} \{\tau\}; \F{esyn} \{\tau\}; \F{rep} \{\tau\}\\

@@

e & ::= & \evar{x} \pipe \elam{\evar{x}}{\tau}{e} \pipe I \pipe I : \tau \pipe I :: \fvar{tycon} \pipe E\\
I & ::= &  \FF{intro}[\tauidx](\splat{e}{1}{n})\\
E & ::= & e\cdot\FF{elim}[\tauidx](\splat{e}{1}{n})\\

@@

 \tau 	& ::= 	& 	\tvar{t} \pipe 
														\tlam{t}{\kappa}{\tau} \pipe 
														\tapp{\tau_1}{\tau_2}  
\\&\pipe&											
														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}

@@

 		& \pipe	& 	  \ell \pipe 
		\tunit \pipe 
														\tpair{\tau_{1}}{\tau_{2}} \pipe 
														\tfst{\tau} \pipe 
														\tsnd{\tau} \pipe \cdots 

@@

%     & \pipe & \tinl{\kappa_2}{\tau_1} \pipe \tinr{\kappa_1}{\tau_2} \\
%     &\pipe& \tsumcase{\tau}{t}{\tau_1}{t}{\tau_2} \\
&\pipe	& 	\ttype{tycon}{\tauidx} \\&\pipe& \tfamcase{\tau}{tycon}{x}{\tau_1}{\tau_2}\\
&\pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
%														\\													%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
\\ 		 & 	\pipe	&	\tden{\tau_2}{\tau_1} \pipe \terr %\pipe \tden{\ibar}{\tau}^{\checkmark} 
\pipe \ttypeof{\tau} \pipe \itransof{\tau} \\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
% %& & \pipe & 

>>>>

%\documentclass[12pt]{article}
\documentclass[10pt,preprint]{sigplanconf}
\newcommand{\lamAce}{\lambda_{\text{Ace}}}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
    \makeatletter
%
% \btIfInRange{number}{range list}{TRUE}{FALSE}
%
% Test if int number <number> is element of a (comma separated) list of ranges
% (such as: {1,3-5,7,10-12,14}) and processes <TRUE> or <FALSE> respectively
%
        \newcount\bt@rangea
        \newcount\bt@rangeb

@@

def log_transfer(t):
  print ("Transferring %s to %s." % (
    __ace_0.decimal_to_str(t[0],2), t[1][0]))

@@

\section{@$\lambda$: Foundations of Active Typechecking and Translation}\label{theory}
\newcommand{\F}[1]{{\sf #1}~}
\newcommand{\FF}[1]{{\sf #1}}
\newcommand{\Q}{\FF{Arg}}
\renewcommand{\tnil}[1]{[]}
\newcommand{\ltxt}[1]{\ell_{\text{#1}}}
\begin{figure*}[t]
\small
$
\begin{array}{l}
\F{using} \phi_{\texttt{nat}}; \phi_{\texttt{lprod}}\\
\elet{one}{\tvar{s}\langle\tvar{z}\langle\rangle\rangle}{\\
\elet{plus}{(\lambda x.\lambda y.\tvar{natrec}~x~\langle y; \lambda p.\lambda r.\tvar{s}\langle r\rangle\rangle) : \tvar{nat} \rightharpoonup \tvar{nat} \rightharpoonup \tvar{nat}}{\\
\FF{let}~y = \{\ltxt{1}=one, \ltxt{2}=plus~one~one\} : \ttype{lprod}{(\ltxt{1}, \tvar{nat}) :: (\ltxt{2}, \tvar{nat}) :: []}}}~\FF{in}\\
y.\ell_2
\end{array}
$
\caption{\small A program that uses the natural number fragment, $\phi_\texttt{nat}$, defined in Figure \ref{nat-atfrag}, written with the syntactic sugar defined in Figure \ref{desugaring}.}
\label{nat-sugared}
\end{figure*}
\begin{figure*}[t]
\small
\hspace{-5px}
$
\begin{array}{l}
\F{using} \hat\phi_{\texttt{nat}}; \hat\phi_{\texttt{lprod}}\\
\elet{one}{\FF{intro}[\FF{in}[\ltxt{s}](\tunit)](\FF{intro}[\FF{in}[\ltxt{z}](\tunit)]()) : \ttype{nat}{\tunit}}{\\
\elet{plus}{(\lambda x.\lambda y.x\cdot\FF{elim}[()](y; \lambda p.\lambda r.\FF{intro}[\FF{in}[\ltxt{s}]()](r)) :\ttype{parr}{(\ttype{nat}{\tunit}, \ttype{parr}{(\ttype{nat}{\tunit}, \ttype{nat}{\tunit})})}}{\\
\FF{let}~y = \FF{intro}[\ell_1 :: \ell_2 :: []](one; plus\cdot\FF{elim}[()](one)\cdot\FF{elim}[()](one)) : \ttype{lprod}{(\ltxt{1}, \tvar{nat}) :: (\ltxt{2}, \tvar{nat}) :: []}}}~\FF{in}\\
y\cdot\FF{elim}[\ell_2]()
\end{array}
$
\caption{\small An equivalent program written in core @$\lambda$ without syntactic sugar and with type-level terms normalized.}
\label{nat-desugared}
\end{figure*}
\begin{figure*}
\small
$
\begin{array}{lcl}
\phi_{\texttt{nat}} & := & \F{tycon}\fvar{nat}~\F{of}\kunit~\{\delta_{\texttt{nat}}\}\\
&& \F{def}\tvar{nat}:\kTypeBlur = \ttype{nat}{\tunit};\\
&& \F{def}\tvar{z}:\Pi[\{\ltxt{idx}\hookrightarrow\kappa_{\text{nat-intro-idx}},\ltxt{ty}\hookrightarrow\kTypeBlur\}]={(\ltxt{idx}=\FF{in}[\ltxt{z}](\tunit), \ltxt{ty}=\tvar{nat})};\\
&& \F{def}\tvar{s}:\Pi[\{\ltxt{idx}\hookrightarrow\kappa_{\text{nat-intro-idx}},\ltxt{ty}\hookrightarrow\kTypeBlur\}]={(\ltxt{idx}=\FF{in}[\ltxt{s}](\tunit), \ltxt{ty}=\tvar{nat})};\\
&& \F{def}\tvar{natrec}:\kappa_{\text{nat-elim-idx}} =\tunit\\
\kappa_{\text{nat-intro-idx}} & := & \Sigma[\{l_z \hookrightarrow \kunit, l_s \hookrightarrow \kunit\}]\\
\kappa_{\text{nat-elim-idx}} & := & \kunit\\
\end{array}
$
%\vspace{-10pt}
\caption{\small The natural number fragment, including definitions used by the assisted intro and elim desugarings, defined and shown being used above.}
\label{nat-atfrag}
\end{figure*}
In this section, we will describe the foundations of \emph{active typechecking and translation} by constructing a core lambda calculus, @$\lambda$. 
A \emph{program} in @$\lambda$, $\rho$, consists of a series of fragment definitions, $\phi$, for use by an external term, $e$. An example of an @$\lambda$ program, $\rho_{\text{example}}$, that uses {fragment definitions} $\phi_{\texttt{nat}}$ and $\phi_{\texttt{lprod}}$, defining primitive natural numbers and labeled products (i.e. ordered records), is shown in Fig. \ref{nat-sugared}. In Ace,  fragment definitions are packaged separately; we do not include this in the core calculus for simplicity.

In this example, we see the introduction and elimination forms for natural numbers, functions and labeled tuples all being used. However, in the core syntax, shown (from the perspective of fragment clients) in Fig. \ref{client-syntax}, there are only two external introductory forms, $\lambda x.e$ and $\FF{intro}[\sigma](\overline{e})$, and one external elimination form, $e\cdot\FF{elim}[\sigma](\overline{e})$, where $\overline{e}$ is shorthand for zero or more semicolon-separated external terms, called the \emph{arguments} of the introduction or elimination operation. The forms used in Fig. \ref{nat-sugared} are \emph{derived forms}, defined in terms of these core forms in Fig. \ref{derived}.

Both fragments and external terms can contain \emph{static terms}, $\sigma$, which are evaluated statically to \emph{static values}, $\hat\sigma$. The static language is itself a simply-typed lambda calculus. We call the types of static terms \emph{kinds}, $\kappa$, by convention. Static variables are written in bold font, $\tvar{x}$, to emphasize that they are distinct from variables bound by external terms, written $x$. For example, in the example in Figure \ref{nat-sugared}, $\tvar{nat}$ is a static variable bound by the fragment $\phi_{\texttt{nat}}$ to the natural number type. Types have kind $\kTypeBlur$ and are introduced by naming a \emph{type constructor}, written in small-caps, e.g. $\fvar{nat}$ and $\fvar{lprod}$, and providing an \emph{index}, which is a static term of the \emph{index kind} of the type constructor.  For example, $\fvar{nat}$ is indexed by $\kunit$ (because there is only one natural number type), so $\tvar{nat}$ is defined to be $\ttype{nat}{\tunit}$, and $\fvar{lprod}$ is indexed by $\klist{\kpair{\FF{L}}{\kTypeBlur}}$, classifying static lists pairing \emph{labels} with types. Labels are static values of kind $\FF{L}$, written abstractly using the metavariables containing $\ell$ in our examples.

Before typechecking the external term in a program, the program is kind checked and its static terms are normalized to static values.  We write static values using the hatted metavariable $\hat \sigma$. Other syntactic metavariables in our system also have hatted forms if they can contain static terms. The desugared, statically normalized version of the example in Figure \ref{nat-sugared}, $\hat\rho_{\text{example}}$, is shown in Figure \ref{nat-normal}. %The syntax of @$\lambda$ from the perspective of a fragment client is given in Fig. \ref{atlam-syntax}. We will see a few additions relevant only to fragment providers when we discuss defining fragments below.

\subsection{External Terms}
 a program consists of first \emph{kind checking} its static terms, then normalizing its static terms, then typechecking the external term and,  simultaneously, \emph{translating} it to a term, $\iota$, in the \emph{typed internal language}. The dynamic behavior of an external term is determined entirely by its translation to this language. Internal types are written $\tau$. The internal language is only exposed to fragment providers, not to clients (i.e. normal developers).

%We will now give a core typed lambda calculus, @$\lambda$, that captures the semantics described in the previous sections. It is intended to make precise how active typechecking and translation works and how our mechanism relates to existing work on bidirectional typechecking, type-level computation and typed compilation while abstracting away from the details of Python's syntax and imposing a stronger type-level semantics that will allow us to state metatheoretic properties of interest. We will assume a fixed base for functions (providing the standard semantics of lambda functions) and target language, which we here call the \emph{internal language}.
 % program written in core @$\lambda$, without syntactic sugar and with all static terms normalized, is shown in Fig. \ref{nat-desugared}. The syntax of core @$\lambda$ is shown in Fig. \ref{grammar} and the syntactic desugarings are shown in Fig. \ref{desugaring}. The definition of $\phi_{\texttt{nat}}$ is shown in Fig. \ref{nat-atfrag}. We will discuss these in the following sections.
%$$\begin{array}{rrcl}	
%%\textbf{programs} & 
%\text{programs} & \rho & ::= & \F{using}\phi~\F{in}e \\
%\text{fragments} & \phi &  ::= & \F{tycon}\fvar{tycon}~\F{of}\kappat{tyidx}~{\{}\F{iana} \tau_1; \F{esyn} \tau_2; \F{rep} \tau_3{\}} %\pipe \phi; \phi %\\
%%\\
%%& \pipe & 
%\pipe \F{def}\tvar{t} : \kappa = \tau \pipe \phi; \phi%\pdef{t}{\kappa}{\tau}{\progsort} \pipe 
%\\ %\pfam{\familyDf}{\progsort}
%%\\&  \pipe & \pdef{t}{\kappa}{\tau}{\progsort}  \pipe e
%%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
%\text{external terms}& e & ::= & \evar{x} \pipe \F{let} x = e_1~\F{in} e_2 \pipe \lambda x.e \pipe e : \tau\\
%&& \pipe & \FF{intro}[\taut{opidx}](\splat{e}{1}{n})\pipe e\cdot\FF{elim}[\taut{opidx}](\splat{e}{1}{n}) \\
%\text{type-level terms} & \tau 	& ::= 	& 	\tvar{t} \pipe \tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} \pipe \terr  \\
% && \pipe & 
% %\F{let}\tvar{t}{:}\kappa=\tau_1~\F{in} \tau_2 \pipe 
%														\tlam{t}{\kappa}{\tau} \pipe 
%														\tapp{\tau_1}{\tau_2} \\
%&&\pipe&											
%														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
%									                     \tfold{\tau_1}{\tau_2}{\tvar{t}_{hd}}{\tvar{t}_{tl}}{\tvar{t}_{rec}}{\tau_3}
% 		&& \pipe	& 	  \ell  \\
%	    && \pipe &  
%		\tunit \pipe 
%														\tpair{\tau_{1}}{\tau_{2}} \pipe 
%														\tfst{\tau} \pipe 
%														\tsnd{\tau} 
%     && \pipe & \tinl{\kappa_1, \kappa_2}{\tau_1} \pipe \tinr{\kappa_1, \kappa_2}{\tau_2} \pipe \tsumcase{\tau}{t}{\tau_1}{t}{\tau_2}\\
%&&\pipe	& 	\ttype{tycon}{\taut{tyidx}} \pipe  \tfamcase{\tau}{tycon}{t}{\tau_1}{\tau_2}\\
%%														\\													%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
%%\\ 		 & 	\pipe	&	\tden{\tau_2}{\tau_1} \pipe \terr %\pipe \tden{\ibar}{\tau}^{\checkmark} 
%%\pipe \ttypeof{\tau} \pipe \itransof{\tau} \\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
%% %& & \pipe & 
%%%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
%%%														 \\
% && \pipe & \titerm{\iota}
%			\pipe	\titype{\sigma} \pipe \trepof{\tau} \\
%		&& \pipe &  (\Gamma; e)? \pipe \FF{syn}(\tau_1; \tvar{t}_{ty},\tvar{t}_{trans}.\tau_2) \pipe \FF{ana}(\tau_1; \tau_2; \tvar{t}_{trans}.\tau_3)\\
%%\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
%%%														\efix{x}{\tau}{e} \pipe 
%%														\elam{\evar{x}}{\tau}{e} \pipe 
%%														\eop{tycon}{op}{
%%															\tauidx
%%														}{
%%  												    		\splat{e}{1}{n}
%%														} \\
%%%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
%%\\
%%\\
%%% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
%%% 												\trepof{\tau} \pipe \dup{\tau}\\
%%\text{translational IL}	& \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 
%%	%\pipe \ilam{x}{\bar \sigma}{\bar \iota} \pipe \iapp{\bar \iota_1}{\bar \iota_2} 
%%	\pipe \cdots \pipe \itransof{\tau} \\						
%% & \bar{\sigma} & ::= & \darrow{\bar \sigma_1}{\bar \sigma_2} \pipe \cdots \pipe \dup{\tau} \pipe \trepof{\tau} \\
%%%\text{abstracted} & \sabs & ::= & \darrow{\sabs_1}{\sabs_2} \pipe \cdots \pipe \sabsrep{\tau}
%\text{kinds} & \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \FF{L} \pipe \kunit \pipe \kpair{\kappa_1}{\kappa_2} \pipe \ksum{\kappa_1}{\kappa_2} \pipe \kTypeBlur \pipe \kITerm \pipe \kIType \pipe \Q
%%											    \klabel \pipe
%%											    \klist{\kappa} \pipe
%%												\kunit \pipe 
%%												\kpair{\kappa_{1}}{\kappa_{2}} \\
%%												&&\pipe&
%%												\ksum{\kappa_1}{\kappa_2} \pipe
%%												\kTypeBlur \pipe \kDen \pipe 
%%												\kIType								
%%%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
%%%											 							&		&		&	\\
%\text{typing contexts} & \Gamma & ::= & \emptyset \pipe \Gamma, x \Rightarrow \tau\\
%\text{internal terms} & \iota	&	::=	&	\evar{x} \pipe 
%												\ifix{\evar{x}}{\sigma}{\iota} \pipe
%												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
%												\iapp{\iota_{1}}{\iota_{2}} \pipe n \pipe \iota_1 \pm \iota_2 \pipe \FF{if0}(\iota; \iota_1; \iota_2)\\
%		&& \pipe &  \iunit \pipe \ipair{\iota_1}{\iota_2} \pipe \ifst{\iota} \pipe \isnd{\iota} \pipe \iup{\tau} 
%%\\&\pipe&								\cdots \pipe 	
%%												n \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}} \\
%%& \pipe & 
%%												\iunit \pipe
%%												\ipair{\iota_{1}}{\iota_{2}} \pipe 
%%												\ifst{\iota} \pipe
%%												\isnd{\iota} 
%%\\&\pipe&												
%%												 \iinl{\sigma_2}{\iota_1} \pipe \iinr{\sigma_1}{\iota_2} \\&\pipe & \icase{e}{x}{e_1}{x}{e_2} \\
%\\	\text{internal types} & \sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} %\pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} \pipe \dsum{\sigma_1}{\sigma_2}  
% \pipe \dup{\tau}
%\end{array}$$
%%\vspace{-10pt}
%\caption{\small Syntax of Core @$\lambda$. Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{tycon}$ ranges over type constructor names, $\ell$ ranges over labels, $n$ ranges over integers and $\pm$ ranges over standard binary operations over integers. The introductory form for arguments, $(\Gamma; e)?$, should only be constructed by the compiler, not by type constructor providers.
%\label{grammar}}
\begin{figure*}[t]

@@

\begin{tabular}{r l l l}
 & \textbf{initial} & \textbf{statically normal}\\
\textbf{programs} & $\rho ::=$ & $\hat\rho ::=$\\
 & ~~$\FF{using}~{\phi}~\FF{in}~e$ & ~~$\FF{using}~{\hat\phi}~\FF{in}~\hat e$\\
\textbf{fragment decls.} & $\phi ::=$ & $\hat\phi ::=$\\
tycon decls. & ~~$\FF{tycon}~\fvar{tycon} : \Theta = \theta$ & ~~$\FF{tycon}~\fvar{tycon} : \Theta = \hat\theta$\\
static binding & ~~$\FF{def}~\tvar{x} : \kappa = \sigma$ & \\
 & ~~$\phi; \phi$ & ~~$\hat\phi; \hat\phi$\\
\textbf{static terms} & $\sigma ::=$ & $\hat\sigma ::= $ &\textbf{kinds}~~~  $\kappa ::= $\\
static values & ~~$\hat \sigma$\\
static variables & ~~$\tvar{x}$ & ~ & ~\\
types & ~~$\ttype{tycon}{\sigma}$ & ~~$\ttype{tycon}{\hat\sigma}$ & ~~$\kTypeBlur$\\
static products & ~~$\tunit$ & ~~$\tunit$ & ~~$\kunit$\\
& ~~$(\sigma, \sigma)$ & ~~$(\hat\sigma, \hat\sigma)$ & ~~$\kpair{\kappa}{\kappa}$\\
static sums & ~~$\FF{in}[\ell](\sigma)$ & ~~$\FF{in}[\ell](\hat\sigma)$ & ~~$\Sigma[\{\overline{\ell \hookrightarrow \kappa}\}]$\\
static lists & ~~$[]_\kappa$ & ~~$[]_\kappa$ & ~~$\klist{\kappa}$\\
 & ~~$\sigma :: \sigma$ & ~~$\hat\sigma :: \hat\sigma$ & \\
static labels & ~~$\ell$ & ~~$\ell$ & ~~$\FF{L}$\\

@@

\textbf{external terms} & $e ::= $ & $\hat e ::=$ \\
variables & ~~$x$ & ~~$x$\\
binding & ~~$\FF{let}~x=e~\FF{in}~e$ & ~~$\FF{let}~x=\hat{e}~\FF{in}~\hat{e}$\\
fixpoints & ~~$\FF{fix}~x{:}\sigma~\FF{is}~e$ & ~~$\FF{fix}~x{:}\hat\sigma~\FF{is}~\hat e$\\
ascription & ~~$e : \sigma$ & ~~$\hat{e} : \hat\sigma$\\
%introductory forms & ~~$\nu$ & ~~$\hat\nu$\\
%eliminatory forms & ~~$\eta$ & ~~$\hat\eta$\\
% & $\nu ::= $ & $\hat \nu ::=$\\
analytic lambda & ~~$\lambda x.e$ & ~~$\lambda x.\hat{e}$\\
%synthetic lambda & ~~$\lambda x {:} \sigma.e$ & ~~$\lambda x{:}\hat\sigma.\hat e$\\
analytic intro & ~~$\FF{intro}[\sigma](\overline{e})$ & ~~$\FF{intro}[\hat\sigma](\overline{\hat e})$\\
%synthetic intro & ~~$\fvar{tycon}\cdot\FF{intro}[\sigma](\overline{e})$ & ~~$\fvar{tycon}\cdot\FF{intro}[\hat \sigma](\overline{\hat e})$\\
% & $\eta ::= $ & $\hat\eta ::=$\\
synthetic elim & ~~$e\cdot\FF{elim}[\sigma](\overline{e})$ & ~~$\hat e\cdot\FF{elim}[\hat \sigma](\overline{\hat e})$\\
\end{tabular}
\caption{Syntax from the perspective of fragment clients (i.e. normal developers)}
\end{figure*}
\begin{figure*}[t]
\hspace{-5px}
$
\begin{array}{rclr}
%\FF{intro}(e_1; \ldots; e_n) & := & \FF{intro}[()](e_1; \ldots; e_n)\\
\sigma\langle e_1; \ldots; e_n \rangle & := & \FF{intro}[\tfst{\sigma}](e_1; \ldots; e_n) : \tsnd{\sigma} & \text{assisted intro}\\
%\langle\sigmaidx\rangle(e_1; \ldots; e_n) & := & \FF{intro}[\sigmaidx](e_1; \ldots; e_n)\\
%[\sigmaidx](e_1; \ldots; e_n) & := & \FF{intro}[\sigmaidx](e_1; \ldots; e_n)\\
\{ \ell_1{=}~e_1, \ldots, \ell_n{=}~e_n \} & := & \FF{intro}[\ell_1 :: \ldots :: \ell_n :: []_{\FF{L}}](e_1; \ldots; e_n) & \text{labeled seq.}  \\
(e_1, ..., e_n) & := & \FF{intro}[()](e_1; \ldots; e_n) & \text{unlabeled seq.}\\
%n & := & \FF{intro}[n]() & \text{number literal} \\
%s & := & \FF{intro}[s]() & \text{string literal}\\
%e\cdot\FF{elim}(e_1; \ldots; e_n) & := & e\cdot\FF{elim}[()](e_1; \ldots; e_n)\\
e~e_1 & := & e\cdot\FF{elim}[()](e_1) & \text{application}\\
\sigma~e~\langle e_1; \ldots; e_n \rangle  & := & e\cdot\FF{elim}[\sigma](e_1; \ldots; e_n) & \text{assisted elim}\\
e.\ell & := & e\cdot\FF{elim}[\ell]() & \text{projection}\\
\F{case}e~\{\ell_1\langle x_1 \rangle \Rightarrow e_1~|~\ldots~|~\ell_n\langle x_n \rangle \Rightarrow e_n\} & := & e\cdot\FF{elim}[\ell_1 :: \ldots :: \ell_n :: []_{\FF{L}}]( & \text{case analysis}\\
 & & \quad \lambda x_1.e_1; \ldots; \lambda x_n.e_n)\\
\sigma_1 \rightharpoonup \sigma_2 & := & \ttype{parr}{(\sigma_1, \sigma_2)} & \text{arrow types}
%!TEX encoding = UTF-8 Unicodee.\ell(e_1; ...; e_n) & := & e\cdot\FF{elim}[\ell](e_1; \ldots; e_n)
%(e_1, ..., e_n) & := & intro[()](e_1; ...; e_n)
%n &  := & intro[n]()
%s & := & intro[s]()
%e.l & := & elim[l](e)
%e[e_1; ...; e_n] & := & elim[()](e_1, ..., e_n)
%e.l(e_1; ...; e_n) & := & elim[l]
\end{array}
$
\caption{\small Desugaring from conventional concrete syntax to core forms. The number and string literal forms assume type-level numbers, $n$, and strings, $s$ (details not shown).}
\label{desugaring}
\end{figure*}
\newcommand{\cev}[1]{\reflectbox{\ensuremath{\vec{\reflectbox{\ensuremath{#1}}}}}}
\newcommand{\tri}{\cev\iota}
\newcommand{\trt}{\cev\tau}
\begin{figure*}
\newcommand{\tcl}[1]{\multicolumn{2}{l}{#1}}
\begin{tabular}{r l l l l}
 & \tcl{\textbf{initial}} & \textbf{statically normal} \\
\textbf{tycon definitions} & \tcl{$\delta ::=$}  & $\hat\delta ::=$ \\
 & \tcl{~~$\FF{iana}~\sigma; \FF{esyn}~\sigma; \FF{rep}~\sigma$} & ~~$\FF{iana}~\hat\sigma; \FF{esyn}~\hat\sigma; \FF{rep}~\hat\sigma$\\

@@

\textbf{(static language, cont.)} & \tcl{$\sigma ::= ...$} & $\hat\sigma ::= ...$ & $\kappa ::= ...$\\
type elim & \tcl{~~$\tfamcase{\sigma}{tycon}{x}{\sigma}{\sigma}$} & \\
static product elim & \tcl{~~$\sigma.\FF{prl}$} &\\
 & \tcl{~~$\sigma.\FF{prr}$} &\\
static sum elim & \tcl{~~$\FF{case}~\sigma~\{\overline{l\langle\tvar{x}\rangle\Rightarrow\sigma}\}$} &\\
static list elim & \tcl{~~$\FF{listrec}(\sigma; \sigma; \tvar{h}, \tvar{t}, \tvar{r}.\sigma)$} & ~\\
decidable equality & \tcl{~~$\tifeq{\hat\sigma}{\hat\sigma}{\kappa}{\hat\sigma}{\hat\sigma}$} & ~ & ~\\
type error & \tcl{~~$\FF{tyerr}$} & ~ \\
static functions & \tcl{~~$\tlam{x}{\kappa}{\sigma}$} & ~~$\tlam{x}{\kappa}{\sigma}$ & ~~$\karrow{\kappa}{\kappa}$\\
 & \tcl{~~$\sigma~\sigma$} & ~ & ~\\
term translations & \tcl{~~$\titerm{\tri}$} & ~~$\titerm{\iota}$ & ~~$\kITerm$\\
type translations & \tcl{~~$\titype{\trt}$} & ~~$\titype{\tau}$ & ~~$\kIType$\\
 & \tcl{~~$\trepof{\sigma}$} & ~\\
arguments & \multicolumn{3}{c}{\color{gray}(introduced only by typechecker)} & ~~$\Q$\\
 & \tcl{~~$\FF{ana}(\sigma; \sigma; \tvar{x}.\sigma)$} & \\
 & \tcl{~~$\FF{syn}(\sigma; \tvar{x}, \tvar{t}.\sigma)$} & \\

@@

\textbf{internal language} & $\tri ::= $ & $\trt ::= $ & $\iota ::= $ & $\tau ::=$\\
splice & ~~$\iup{\sigma}$ & ~~$\dup{\sigma}$ & \\
variables & ~~$x$ &  & ~~$x$ & \\
fixpoints & ~~$\FF{fix}~x{:}\trt~\FF{is}~\tri$ & & ~~$\FF{fix}~x{:}\tau~\FF{is}~\iota$\\
functions & ~~$\lambda x{:}\trt.\tri$ & ~~$\trt \rightharpoonup \trt$ & ~~$\lambda x{:}\tau.\iota$ & ~~$\tau \rightharpoonup \tau$\\
 & ~~$\tri~\tri$ & ~ & ~~$\iota~\iota$\\
products & ~~$()$ & ~~$\dunit$ & ~~$()$ & ~~$\dunit$\\
 & ~~$(\tri, \tri)$ & ~~$\dpair{\trt}{\trt}$ & ~~$(\iota, \iota)$ & ~~$\dpair{\tau}{\tau}$\\
 & ~~$\tri.\FF{prl}$ & ~ & ~~$\tri.\FF{prl}$ & ~\\
  & ~~$\tri.\FF{prr}$ & ~ & ~~$\tri.\FF{prr}$ & ~\\
%products & ~~$(\overline{l=\tri})$ & ~~$\Pi[\overline{(l, \trt)}]$ & ~~$(\overline{l = \iota})$ & ~~$\Pi[\overline{(l, \tau)}]$\\
% & ~~$\tri\cdot l$ & ~ & ~~$\iota\cdot l$ & ~\\
sums & ~~$\FF{in}[\ell](\tri)$ & ~~$\Sigma[\{\overline{\ell \hookrightarrow \trt}\}]$ & ~~$\FF{in}[\ell](\iota)$ & ~~$\Sigma[\{\overline{\ell \hookrightarrow \tau}\}]$\\
 & ~~$\FF{case}~ \tri ~\{\overline{\ell(x) \Rightarrow \tri}\}$ & ~ & ~~$\FF{case}~\iota~\{\overline{\ell(x)\Rightarrow \iota}\}$ & ~\\
integers & ~~$\FF{int}[z]$ & ~~$\FF{int}$ & ~~$\FF{int}[z]$ & ~~$\FF{int}$\\
 & ~~$\tri \pm \tri$ & ~ & ~~$\iota \pm \iota$ & ~\\
 & ~~$\FF{if0}(\tri; \tri; \tri; \tri)$ & ~ & ~~$\FF{if0}(\iota; \iota; \iota; \iota)$ & ~
 %& ~~$
\end{tabular}
\caption{Syntax from the perspective of fragment providers also includes an internal language and static language.}
\end{figure*}
\begin{figure*}
\begin{tabular}{r l l}
external typing context & $\Gamma ::= \emptyset \pipe \Gamma, x \Rightarrow \ttype{tycon}{\hat\sigma}$\\

@@

 & \textbf{initial} & \textbf{statically normal} \\
\textbf{static language, cont.} & $\sigma ::= ...$ & $\hat\sigma ::= ...$\\
argument intro & ~~$\FF{arg}(\Gamma; \hat e)$ & ~~$\FF{arg}(\Gamma; \hat e)$
\end{tabular}
\caption{Syntax from the perspective of the semantics / compiler.}
\end{figure*}
\newcommand{\fkctx}{\mathcal{F}}
\newcommand{\ok}[1]{#1~\mathtt{ok}}
\newcommand{\normalize}[3]{#1 \curlyveedownarrow_{#2} #3}
\newcommand{\normalizeX}[2]{\normalize{#1}{\Phi}{#2}}
\newcommand{\serr}[1]{#1~\mathtt{tyerr}}
\begin{figure*}
\begin{tabular}{r l l}
\textbf{context} & \textbf{syntax} & \textbf{well-formedness}\\
fragment kind context & $\fkctx ::= \fkctx_0 \pipe \fkctx, \fvar{tycon}[\kappa]~\{\FF{intro}[\kappa]; \FF{elim}[\kappa]\}$ & $\vdash \fkctx$\\
